\subsection{Hearing}

\sepslide{Hearing}



\subsubsection{\bh{šå̄maʿ}}

\begin{frame}{\bh{šå̄maʿ} ‘to hear’}
	\begin{center}
		$$
		\mtext{\bh{šå̄maʿ}} \transto
		\begin{dcases}
			\C{\hl{clywed}}\\
			\C{\hl{gwrando}}\\
			\mtext{\small\C{(deall)}}
		\end{dcases}
		$$
	\end{center}
\end{frame}



\begin{frame}{\ex Content (\bh{kī} ‘that’)}
	\begin{example}{Gen.}{42}{2}{}{}
		\quoling
		{וַיֹּ֕אמֶר הִנֵּ֣ה \lhl{שָׁמַ֔עְתִּי} \hlA{כִּ֥י} יֶשׁ־שֶׁ֖בֶר בְּמִצְרָ֑יִם רְדוּ־שָׁ֙מָּה֙ וְשִׁבְרוּ־לָ֣נוּ מִשָּׁ֔ם וְנִחְיֶ֖ה וְלֹ֥א נָמֽוּת׃}
		{Dywedodd hefyd, wele \lhl{clywais} \hlA{fod} ŷd yn yꝛ Aipht, ewch i wared yno, a phꝛynnwch i ni oddi yno, fel y bôm fyw, ac na byddom feirw.}
		{way·yōmɛr hinnē \lhl{šå̄maʿtī} \hlA{kī} yɛš·šɛḇɛr bə·miṣrå̄yim rəḏū·šå̄mm·å̄ wə·šiḇrū·l·å̄nū miš·šå̄m wə·niḥyɛ wə·lō nå̄mūṯ}
		{And he said, Behold, I have \lhl{heard} \hlA{that} there is corn in Egypt: get you down thither, and buy for us from thence; that we may live, and not die.}
	\end{example}
\end{frame}



\begin{frame}{\ex Content (with a form of \bh{ʾå̄mar} ‘to say’)}
	\begin{example}{Gen.}{41}{15}{1}{}
		\quoling
		{וַיֹּ֤אמֶר פַּרְעֹה֙ אֶל־יֹוסֵ֔ף חֲלֹ֣ום חָלַ֔מְתִּי וּפֹתֵ֖ר אֵ֣ין אֹתֹ֑ו וַאֲנִ֗י \lhl{שָׁמַ֤עְתִּי} עָלֶ֙יךָ֙ \hlA{לֵאמֹ֔ר} תִּשְׁמַ֥ע חֲלֹ֖ום לִפְתֹּ֥ר אֹתֹֽו׃}
		{A Pharao a ddywedodd wꝛth Joſeph, bꝛeuddwydiais freuddwyd, ac nid [oes] ai deonglo ef: ond myfi a \lhl{glywais} \hlA{ddywedyd} am danat ti, y gwꝛandewi freuddwyd iw ddeonglu.}
		{way·yōmɛr parʿō ʾɛl·yōsēp̄ ḥălōm ḥå̄lamtī ū·p̄ōṯēr ʾēn ʾōṯ·ō wa·ʾănī \lhl{šå̄maʿtī} ʿå̄lɛḵå̄ \hlA{lē·mōr} tišmaʿ ḥălōm li·p̄tōr ʾōṯ·ō}
		{And Pharaoh said unto Joseph, I have dreamed a dream, and \emph{there is} none that can interpret it: and I have \lhl{heard} say of thee, \hlA{\emph{that}} thou canst understand a dream to interpret it.}
	\end{example}
\end{frame}



\begin{frame}{\ex \bh{šå̄maʿ}~+ \textsc{preposition}}
	\begin{example}{Ex.}{4}{1}{}{}
		\quoling
		{וַיַּ֤עַן מֹשֶׁה֙ וַיֹּ֔אמֶר וְהֵן֙ לֹֽא־יַאֲמִ֣ינוּ לִ֔י וְלֹ֥א \lhl{יִשְׁמְע֖וּ} \hlA{בְּ}קֹלִ֑י כִּ֣י יֹֽאמְר֔וּ לֹֽא־נִרְאָ֥ה אֵלֶ֖יךָ יְהוָֽה׃}
		{Yna Moſes a attebodd, ac a ddywedodd, etto wele ni chꝛedant i mi ac ni \lhl{wꝛandawant} \hlA{ar} fy llais: onid dywedant nid ymgdangoſodd yꝛ Arglwydd i ti.}
		{way·yaʿan mōšɛ way·yōmɛr wə·hēn lō·yaʾămīnū l·ī wə·lō \lhl{yišməʿū} \hlA{bə·}qōlī kī yōmrū lō·nirʾå̄ ʾēlɛḵå̄ {\YHWH}}
		{And Moses answered and said, But, behold, they will not believe me, nor \lhl{hearken} \hlA{unto} my voice: for they will say, The {\LORD} hath not appeared unto thee.}
	\end{example}
\end{frame}



\begin{frame}{\ex Judgement}
	\begin{example}{Deut.}{1}{16}{}{}
		\quoling
		{וָאֲצַוֶּה֙ אֶת־שֹׁ֣פְטֵיכֶ֔ם בָּעֵ֥ת הַהִ֖וא לֵאמֹ֑ר \lhl{שָׁמֹ֤עַ} בֵּין־אֲחֵיכֶם֙ וּשְׁפַטְתֶּ֣ם צֶ֔דֶק בֵּֽין־אִ֥ישׁ וּבֵין־אָחִ֖יו וּבֵ֥ין גֵּרֽוֹ׃}
		{A’r amſer hwnnw y goꝛchymynnais i’ch barnwŷꝛ chwi gan ddywedyd: \lhl{gwꝛandewch} [ddadleuon] rhwng eich bꝛodyꝛ, a bernwch gyfiawnder rhwng gŵꝛ ai frawd, ac ai eſtron hefyd.}
		{wå̄·ʾăṣawwɛ ʾɛṯ·šōp̄ṭēḵɛm b·å̄·ʿēṯ ha·hi lē·mōr \lhl{šå̄mōaʿ} bēn·ʾăḥēḵɛm ū·šp̄aṭtɛm ṣɛḏɛq bēn·ʾīš ū·ḇēn·ʾå̄ḥīw ū·ḇēn gērō}
		{And I charged your judges at that time, saying, \lhl{Hear} \emph{the causes} between your brethren, and judge righteously between \emph{every} man and his brother, and the stranger \emph{that is} with him.}
	\end{example}
\end{frame}


\begin{frame}{\ex Interpretation of dreams}
	\begin{example}{Gen.}{41}{15}{2}{}
		\quoling
		{וַיֹּ֤אמֶר פַּרְעֹה֙ אֶל־יֹוסֵ֔ף חֲלֹ֣ום חָלַ֔מְתִּי וּפֹתֵ֖ר אֵ֣ין אֹתֹ֑ו וַאֲנִ֗י שָׁמַ֤עְתִּי עָלֶ֙יךָ֙ לֵאמֹ֔ר \lhl{תִּשְׁמַ֥ע} \hlA{חֲלֹ֖ום} לִפְתֹּ֥ר אֹתֹֽו׃}
		{A Pharao a ddywedodd wꝛth Joſeph, bꝛeuddwydiais freuddwyd, ac nid [oes] ai deonglo ef: ond myfi a glywais ddywedyd am danat ti, y \lhl{gwꝛandewi} \hlA{freuddwyd} iw ddeonglu.}
		{way·yōmɛr parʿō ʾɛl·yōsēp̄ ḥălōm ḥå̄lamtī ū·p̄ōṯēr ʾēn ʾōṯ·ō wa·ʾănī šå̄maʿtī ʿå̄lɛḵå̄ lē·mōr \lhl{tišmaʿ} \hlA{ḥălōm} li·p̄tōr ʾōṯ·ō}
		{And Pharaoh said unto Joseph, I have dreamed a dream, and \textit{there is} none that can interpret it: and I have heard say of thee, \textit{that} thou canst \lhl{understand} a \hlA{dream} to interpret it.}
	\end{example}
\end{frame}



\begin{frame}{\ex Interpretation of dreams}
	\begin{example}{Gen.}{37}{6}{}{}
		\quoling
		{וַיֹּ֖אמֶר אֲלֵיהֶ֑ם \lhl{שִׁמְעוּ}־נָ֕א \hlA{הַחֲלֹ֥ום} הַזֶּ֖ה אֲשֶׁ֥ר חָלָֽמְתִּי׃}
		{O blegit dywedaſe wꝛthynt, \lhl{gwꝛandewch} atolwg y \hlA{bꝛeuddwyd} hwn, yꝛ hwn a frenddwydiais.}
		{way·yōmɛr ʾălēhɛm \lhl{šimʿū}·nå̄ ha·\hlA{ḥălōm} haz·zɛ ʾăšɛr ḥå̄lå̄mtī}
		{And he said unto them, \lhl{Hear}, I pray you, this \hlA{dream} which I have dreamed:}
	\end{example}
\end{frame}



\begin{frame}{\ex A dream’s narrative or interpretation}
	\begin{example}{Judg.}{7}{15}{}{}
		\quoling
		{וַיְהִי֩ \lhl{כִשְׁמֹ֨עַ} גִּדְע֜וֹן אֶת־\hlA{מִסְפַּ֧ר הַחֲל֛וֹם} וְאֶת־\hlA{שִׁבְר֖וֹ} וַיִּשְׁתָּ֑חוּ […]}% וַיָּ֙שָׁב֙ אֶל־מַחֲנֵ֣ה יִשְׂרָאֵ֔ל וַיֹּ֣אמֶר ק֔וּמוּ כִּֽי־נָתַ֧ן יְהוָ֛ה בְּיֶדְכֶ֖ם אֶת־מַחֲנֵ֥ה מִדְיָֽן׃}
		{Pan \lhl{glybu} Gedeon \hlA{adꝛoddiad y bꝛeuddwyd} ai \hlA{ddirnad}, yna efe a addolodd: […]} %ac a ddychwelodd i werſſyll Iſrael, ac a ddywedodd, cyfodwch, canys rhoddodd yꝛ Arglwydd werſſyll y Madianiaid yn eich llaw chwi.}
		{way·hī \lhl{ḵi·šmōaʿ} giḏʿōn ʾɛṯ·\hlA{mispar ha·ḥălōm} wə·ʾɛṯ·\hlA{šiḇrō} way·yištå̄ḥū […]}% way·yå̄šå̄ḇ ʾɛl·maḥănē yiśrå̄ʾēl way·yōmɛr qūmū kī·nå̄ṯan {\YHWH} bə·yɛḏḵɛm ʾɛṯ·maḥănē miḏyå̄n}
		{And it was \textit{so}, when Gideon \lhl{heard} the \hlA{telling of the dream}, and the \hlA{interpretation} thereof, that he worshipped, […]}% and returned into the host of Israel, and said, Arise; for the {\LORD} hath delivered into your hand the host of Midian.}
	\end{example}
\end{frame}



\subsubsection{\bh{hɛʾɛ̆zīn}}

\subsubsection{\bh{hiṭṭå̄}~+ \bh{ʾōzɛn}}

\subsubsection{Minor verbs}

%\paragraph{På̄ṯaḥ}
