\subsection{Sight}

\sepslide{Sight}



\subsubsection{\bh{rå̄ʾå̄}}

\begin{frame}{\bh{rå̄ʾå̄} ‘to see’}
	\begin{center}
		$$
		\mtext{\bh{rå̄ʾå̄}} \transto
		\begin{dcases}
			\C{\hl{gweled}}\\
			\C{\hl{edrych}}\\
			\mtext{\small\C{(canfod)}}
		\end{dcases}
		$$

		$$
		\mtext{\bh{hɛrʾå̄} \gram{(caus.)}} \transto
		\begin{dcases}
			\C{\hl{dangos}}\\
			\mtext{\small\C{(peri + gweled)}}
		\end{dcases}
		$$

		$$
		\mtext{\bh{nirʾå̄} \gram{(pass.)}} \transto
		\begin{dcases}
			\C{\hl{ymddangos}}\\
			\mtext{\small\C{(gweled \gram{(pass.))}}}
		\end{dcases}
		$$
	\end{center}
\end{frame}


\begin{frame}{\ex Sensory perception}
	\begin{example}{Ex.}{10}{23}{}{}
		\quoling
		{לֹֽא־\lhl{רָא֞וּ} אִ֣ישׁ אֶת־אָחִ֗יו וְלֹא־קָ֛מוּ אִ֥ישׁ מִתַּחְתָּ֖יו שְׁלֹ֣שֶׁת יָמִ֑ים וּֽלְכָל־בְּנֵ֧י יִשְׂרָאֵ֛ל הָ֥יָה א֖וֹר בְּמוֹשְׁבֹתָֽם׃}
		{Ni \lhl{wele} neb ei gilydd, ac ni chododd neb oi le dꝛi diwꝛnod: ond yꝛ ydoedd goleuni i holl feibion Iſrael yn eu trigfannau.}
		{lō·\lhl{rå̄ʾū} ʾīš ʾɛṯ·ʾå̄ḥīw wə·lō·qå̄mū ʾīš mit·taḥtå̄w šəlōšɛṯ yå̄mīm ū·l·ḵå̄l·bənē yiśrå̄ʾēl hå̄yå̄ ʾōr bə·mōšḇōṯå̄m}
		{They \lhl{saw} not one another, neither rose any from his place for three days: but all the children of Israel had light in their dwellings.}
	\end{example}
\end{frame}


\begin{frame}{\ex Sensory perception}
	\begin{example}{Num.}{13}{33}{}{}
		\quoling
		{וְשָׁ֣ם \lhl{רָאִ֗ינוּ} אֶת־הַנְּפִילִ֛ים בְּנֵ֥י עֲנָ֖ק מִן־הַנְּפִלִ֑ים וַנְּהִ֤י בְעֵינֵ֙ינוּ֙ כַּֽחֲגָבִ֔ים וְכֵ֥ן הָיִ֖ינוּ בְּעֵינֵיהֶֽם׃}
		{Ac yno y \lhl{gwelſom} feibion Anac y cawꝛi [y rhai a ddaethant] o’ꝛ cawꝛi, ac yꝛ oeddem yn ein golwg ein hunain fel ceiliogod rhedyn, ac felly yꝛ oeddem yn eu golwg hwyntau.}
		{wə·šå̄m \lhl{rå̄ʾīnū} ʾɛṯ·han·nəp̄īlīm bənē ʿănå̄q min·han·nəp̄īlīm wan·nəhī ḇə·ʿēnēnū ka·ḥăḡå̄ḇīm wə·ḵēn hå̄yīnū bə·ʿēnēhɛm}
		{And there we \lhl{saw} the giants, the sons of Anak, \emph{which come} of the giants: and we were in our own sight as grasshoppers, and so we were in their sight.}
	\end{example}
\end{frame}


\begin{frame}{\ex Content (\bh{kī} ‘that’)}
	\begin{example}{Gen.}{1}{4}{}{}
		\quoling
		{\lhl{וַיַּ֧רְא} אֱלֹהִ֛ים אֶת־הָא֖וֹר \hlA{כִּי}־ט֑וֹב וַיַּבְדֵּ֣ל אֱלֹהִ֔ים בֵּ֥ין הָא֖וֹר וּבֵ֥ין הַחֹֽשֶׁךְ׃}
		{Yna Duw a \lhl{welodd} y goleuni \hlA{mai} dâ [oedd,] a Duw a wahanodd rhwng y goleuni a’r tywyllwch.}
		{way·\lhl{yar} ʾɛ̆lōhīm ʾɛṯ·hå̄·ʾōr \hlA{kī}·ṭōḇ way·yaḇdēl ʾɛ̆lōhīm bēn hå̄·ʾōr ū·ḇēn ha·ḥōšɛḵ}
		{And God saw the light, \hlA{that} \emph{it was} good: and God divided the light from the darkness.}
	\end{example}
\end{frame}


\begin{frame}{\ex Content (\bh{kī} ‘that’)}
	\begin{example}{Gen.}{37}{4}{}{}
		\quoling
		{\lhl{וַיִּרְא֣וּ} אֶחָ֗יו \hlA{כִּֽי}־אֹת֞וֹ אָהַ֤ב אֲבִיהֶם֙ מִכָּל־אֶחָ֔יו וַֽיִּשְׂנְא֖וּ אֹת֑וֹ וְלֹ֥א יָכְל֖וּ דַּבְּר֥וֹ לְשָׁלֹֽם׃}
		{Pan \lhl{welodd} ei frodyꝛ, \hlA{fod} eu tâd \hlA{yn} ei garu ef yn fwy nai holl frodyꝛ: yna hwy ai caſaſant ef, ac ni fedꝛent ymddiddan [ag] ef yn heddychol.}
		{way·\lhl{yirʾū} ʾɛḥå̄w \hlA{kī}·ʾōṯ·ō ʾå̄haḇ ʾăḇīhɛm mik·kå̄l·ʾɛḥå̄w way·yiśnəʾū ʾōṯ·ō wə·lō yå̄ḵlū dabbərō lə·šå̄lōm}
		{And when his brethren \lhl{saw} \hlA{that} their father loved him more than all his brethren, they hated him, and could not speak peaceably unto him.}
	\end{example}
\end{frame}


\begin{frame}{\ex Inspection}
	\begin{example}{Num.}{13}{18}{}{}
		\quoling
		{\lhl{וּרְאִיתֶ֥ם} אֶת־הָאָ֖רֶץ מַה־הִ֑וא וְאֶת־הָעָם֙ הַיֹּשֵׁ֣ב עָלֶ֔יהָ הֶחָזָ֥ק הוּא֙ הֲרָפֶ֔ה הַמְעַ֥ט ה֖וּא אִם־רָֽב׃}
		{Ac \lhl{edꝛychwch} y wlad beth yw hi, a’r bobl ſydd yn trigo ynddi, ai cryf, ai gwan, ai llawer [ydynt.]}
		{ū·\lhl{rʾīṯɛm} ʾɛṯ·hå̄·ʾå̄rɛṣ ma·hī wə·ʾɛṯ·hå̄·ʿå̄m hay·yōšēḇ ʿå̄lɛhå̄ hɛ·ḥå̄zå̄q hū hă·rå̄p̄ɛ ha·mʿaṭ hū ʾim·rå̄ḇ}
		{And \lhl{see} the land, what it \kjvit{is}; and the people that dwelleth therein, whether they \kjvit{be} strong or weak, few or many;}
	\end{example}
\end{frame}


\begin{frame}{\ex Inspection}
	\begin{example}{Num.}{32}{8}{}{}
		\quoling
		{כֹּ֥ה עָשׂ֖וּ אֲבֹתֵיכֶ֑ם בְּשָׁלְחִ֥י אֹתָ֛ם מִקָּדֵ֥שׁ בַּרְנֵ֖עַ \lhl{לִרְא֥וֹת} אֶת־הָאָֽרֶץ׃}
		{Felly y gwnaeth eich tadau, pan anfonais hwynt o Cades Barnea i \lhl{edꝛych} y tîr.}
		{kō ʿå̄śū ʾăḇōṯēḵɛm bə·šå̄lḥī ʾōṯå̄·m miq·qå̄ḏēš barnēaʿ li·\lhl{rʾōṯ} ʾɛṯ·hå̄·ʾå̄rɛṣ}
		{Thus did your fathers, when I sent them from Kadeshbarnea to \lhl{see} the land.}
	\end{example}
\end{frame}


\begin{frame}{\ex Priestly examination of leprosy}
	\begin{example}{Lev.}{13}{8}{}{}
		\quoling
		{\lhl{וְרָאָה֙} הַכֹּהֵ֔ן וְהִנֵּ֛ה פָּשְׂתָ֥ה הַמִּסְפַּ֖חַת בָּע֑וֹר וְטִמְּא֥וֹ הַכֹּהֵ֖ן צָרַ֥עַת הִֽוא׃}
		{Ac \lhl{edꝛyched} yꝛ offeiriad, ac os lledodd y grammen yn y croen, yna barned yꝛ offeiriad ef yn aflan: gwahan-glwyf yw hwnnw.}
		{wə·\lhl{rå̄ʾå̄} hak·kōhēn wə·hinnē på̄śṯå̄ ham·mispaḥaṯ b·å̄·ʿōr wə·ṭimməʾō hak·kōhēn ṣå̄raʿaṯ hī}
		{And \emph{if} the priest \lhl{see} that, behold, the scab spreadeth in the skin, then the priest shall pronounce him unclean: it \emph{is} a leprosy.}
	\end{example}
\end{frame}



\subsubsection{\bh{på̄nå̄}}

\begin{frame}{\bh{på̄nå̄} ‘to turn, to face, to look’}
	\begin{center}
		$$
		\mtext{\bh{på̄nå̄}} \transto
		\begin{dcases}
			\C{\hl{edrych}}\\
			\C{\hl{troi}}\\
			\mtext{\small\C{(dychwelyd, wynebu, …)}}
		\end{dcases}
		$$
	\end{center}
\end{frame}


\begin{frame}{\bh{på̄nå̄} being translated by \C{edrych}}
	\begin{itemize}
		\item looking\hfill\exvref{Ex.}{2}{12}{}
		\item accepting or willing {\small (with \C{ar} or \C{am})}\hfill\exvref{2 Chr.}{6}{19}{}
		\item a geographical direction {\small (with \C{tua})}\hfill\exvref{Ezek.}{43}{1}{}
	\end{itemize}
\end{frame}


\begin{frame}{\exwref \bh{på̄nå̄} → \C{edrych} (looking)}
	\exstep
	\begin{example}{Ex.}{2}{12}{}{}
		\quoling
		{\lhl{וַיִּ֤פֶן} כֹּה֙ וָכֹ֔ה וַיַּ֖רְא כִּ֣י אֵ֣ין אִ֑ישׁ וַיַּךְ֙ אֶת־הַמִּצְרִ֔י וַֽיִּטְמְנֵ֖הוּ בַּחֽוֹל׃}
		{Ac efe a \lhl{edꝛychodd} ymma, ac accw, a phan welodd nad [oedd yno] neb: yna efe a laddodd yꝛ Aiphtiad, ac ai cuddiodd yn y tyfod.}
		{way·\lhl{yīp̄ɛn} kō wå̄·ḵō way·yar kī ʾēn ʾīš way·yaḵ ʾɛṯ·ham·miṣrī wayyiṭmənēhū b·a·ḥōl}
		{And he \lhl{looked} this way and that way, and when he saw that \kjvit{there was} no man, he slew the Egyptian, and hid him in the sand.}
	\end{example}
\end{frame}


\begin{frame}{\exwref \bh{på̄nå̄} → \C{edrych} (accepting or willing)}
	\exstep
	\begin{example}{2 Chr.}{6}{19}{}{}
		\quoling
		{\lhl{וּפָנִ֜יתָ} \hlA{אֶל}־תְּפִלַּ֧ת עַבְדְּךָ֛ וְאֶל־תְּחִנָּת֖וֹ יְהוָ֣ה אֱלֹהָ֑י לִשְׁמֹ֤עַ אֶל־הָרִנָּה֙ וְאֶל־הַתְּפִלָּ֔ה אֲשֶׁ֥ר עַבְדְּךָ֖ מִתְפַּלֵּ֥ל לְפָנֶֽיךָ׃}
		{\lhl{Edꝛych} gan hynny \hlA{ar} weddi dy wâs, ac ar ei ddeiſyfiad ef ô Arglwydd fy Nuw: i wꝛando ar y llêf, ac ar y weddi yꝛ hon y mae dy wâs yn ei gweddio ger dy fron di.}
		{ū·\lhl{p̄å̄nīṯå̄} \hlA{ʾɛl}·təp̄illaṯ ʿaḇdəḵå̄ wə·ʾɛl·təḥinnå̄ṯō {\YHWH} ʾɛ̆lōhå̄y li·šmōaʿ ʾɛl·hå̄·rinnå̄ wə·ʾɛl·hat·təp̄illå̄ ʾăšɛr ʿaḇdəḵå̄ miṯpallēl lə·p̄å̄nɛḵå̄}
		{\lhl{Have respect} therefore \hlA{to} the prayer of thy servant, and to his supplication, O {\LORD} my God, to hearken unto the cry and the prayer which thy servant prayeth before thee:}
	\end{example}
\end{frame}


\begin{frame}{\exwref \bh{på̄nå̄} → \C{edrych} (geographical direction)}
	\exstep
	\begin{example}{Ezek.}{43}{1}{}{}
		\quoling
		{וַיּוֹלִכֵ֖נִי אֶל־הַשָּׁ֑עַר שַׁ֕עַר אֲשֶׁ֥ר \lhl{פֹּנֶ֖ה} דֶּ֥רֶךְ הַקָּדִֽים׃}
		{Ac efe a’m dug fi i’r poꝛth [ſef] y poꝛth yꝛ hwn ſydd yn \lhl{edꝛych} \hlA{tua} ’r dwyꝛain.}
		{wayyōlīḵēnī ʾɛl·haš·šå̄ʿar šaʿar ʾăšɛr pōnɛ dɛrɛḵ haq·qå̄ḏīm}
		{Afterward he brought me to the gate, \kjvit{even} the gate that looketh toward the east:}
	\end{example}
\end{frame}


\begin{frame}{\bh{på̄nå̄} being translated by \C{troi}}
	\begin{itemize}
		\item physical sense (+motion)\hfill\exvref{Gen.}{18}{22}{}
		\item metaphorically, turning unto other gods\hfill\exvref{Deut.}{31}{18}{}
	\end{itemize}
\end{frame}


\begin{frame}{\exwref \bh{på̄nå̄} → \C{troi} (physical sense)}
	\exstep
	\begin{example}{Gen.}{18}{22}{}{}
		\quoling
		{\lhl{וַיִּפְנ֤וּ} מִשָּׁם֙ הָֽאֲנָשִׁ֔ים וַיֵּלְכ֖וּ סְדֹ֑מָה וְאַ֨בְרָהָ֔ם עוֹדֶ֥נּוּ עֹמֵ֖ד לִפְנֵ֥י יְהוָֽה׃}
		{A [dau] oꝛ gwŷꝛ a \lhl{dꝛoeſant} oddi yno, ac a aethant tua Sodoma, ac Abꝛaham yn ſefyll \typo{fefyll} etto ger bꝛon yꝛ Arglwydd.}
		{way·\lhl{yip̄nū} miš·šå̄m hå̄·ʾănå̄šīm way·yēlḵū səḏōm·å̄ wə·ʾaḇrå̄hå̄m ʿōḏɛnnū ʿōmēḏ li·p̄nē {\YHWH}}
		{And the men \lhl{turned their faces} from thence, and went toward Sodom: but Abraham stood yet before the {\LORD}.}
	\end{example}
\end{frame}


\begin{frame}{\exwref \bh{på̄nå̄} → \C{troi} (metaphorically)}
	\exstep
	\begin{example}{Deut.}{31}{18}{}{}
		\quoling
		{וְאָנֹכִ֗י הַסְתֵּ֨ר אַסְתִּ֤יר פָּנַי֙ בַּיּ֣וֹם הַה֔וּא עַ֥ל כָּל־הָרָעָ֖ה אֲשֶׁ֣ר עָשָׂ֑ה כִּ֣י \lhl{פָנָ֔ה} אֶל־אֱלֹהִ֖ים אֲחֵרִֽים׃}
		{Canys myfi gan guddio a guddiaf fy wyneb y dydd hwnnw, am yꝛ holl ddꝛygioni ’r hyn a wnaeth efe, pan \lhl{dꝛôdd} at dduwiau dieithꝛ.}
		{wə·ʾå̄nōḵī hastēr ʾastīr på̄nay b·ay·yōm ha·hū ʿal kå̄l·hå̄·rå̄ʿå̄ ʾăšɛr ʿå̄śå̄ kī \lhl{p̄å̄nå̄} ʾɛl·ʾɛ̆lōhīm ʾăḥērīm}
		{And I will surely hide my face in that day for all the evils which they shall have wrought, in that they are \lhl{turned} unto other gods.}
	\end{example}
\end{frame}



\subsubsection{\bh{hibbīṭ}}

\begin{frame}{\bh{hibbīṭ} ‘to look’}
	\begin{center}
		$$
		\mtext{\bh{hibbīṭ}} \transto
		\mtext{\C{edrych}}
		$$
	\end{center}
\end{frame}


\begin{frame}{\ex \bh{hibbīṭ} ‘to look’ → \C{edrych}}
	\begin{example}{Prov.}{4}{25}{}{}
		\quoling
		{עֵ֭ינֶיךָ לְנֹ֣כַח \lhl{יַבִּ֑יטוּ} וְ֝עַפְעַפֶּ֗יךָ יַיְשִׁ֥רוּ נֶגְדֶּֽךָ׃}
		{\lhl{Edꝛyched} dy lygaid ar yꝛ iniawn, a chyfarwydded amrantau dy lygaid dy ffoꝛdd o’th flaen.}
		{ʿēnɛḵå̄ lə-nōḵaḥ \lhl{yabbīṭū} wə-ʿap̄ʿappɛḵå̄ yayšīrū nɛḡdɛḵå̄}
		{Let thine eyes \lhl{look} right on, and let thine eyelids look straight before thee.}
	\end{example}
\end{frame}



\subsubsection{\bh{ḥå̄zå̄}}

\begin{frame}{\bh{ḥå̄zå̄} ‘to see, to prophesy’}
	\begin{center}
		$$
		\mtext{\bh{ḥå̄zå̄}} \transto
		\begin{dcases}
			\C{\hl{gweled}}\\
			\mtext{\small\C{(…)}}
		\end{dcases}
		$$
	\end{center}
\end{frame}


\begin{frame}{\ex \bh{ḥå̄zå̄} ‘to see, to prophesy’}
	\begin{example}{Isa.}{2}{1}{}{}
		\quoling
		{הַדָּבָר֙ אֲשֶׁ֣ר \lhl{חָזָ֔ה}	יְשַֽׁעְיָ֖הוּ בֶּן־אָמ֑וֹץ עַל־יְהוּדָ֖ה וִירוּשָׁלִָֽם׃}
		{Y gair yꝛ hwn a \lhl{welodd} Eſay mab Amos am Iuda, ac Ieruſalem.}
		{had·då̄ḇå̄r ʾăšɛr \lhl{ḥå̄zå̄} yəšaʿyå̄hū bɛn·ʾå̄mōṣ ʿal·yəhūḏå̄ wī·rūšå̄lå̄yim}
		{The word that Isaiah the son of Amoz \lhl{saw} concerning Judah and Jerusalem.}
	\end{example}
\end{frame}

\begin{frame}{\ex \bh{rå̄ʾå̄} and \bh{ḥå̄zå̄}, lexical differentiation}
	\begin{example}{Isa.}{30}{10}{}{}
		\quoling
		{אֲשֶׁ֨ר אָמְר֤וּ \hlA{לָֽרֹאִים֙} לֹ֣א \hlA{תִרְא֔וּ} \lhl{וְלַ֣חֹזִ֔ים} לֹ֥א \lhl{תֶחֱזוּ}־לָ֖נוּ נְכֹח֑וֹת דַּבְּרוּ־לָ֣נוּ חֲלָק֔וֹת \lhl{חֲז֖וּ} מַהֲתַלּֽוֹת׃}
		{Y rhai a ddywedaſant wꝛth y rhai a \hlA{welant}, na \hlA{welwch}, ac wꝛth y rhai a \lhl{ganfyddant}, na \lhl{chanfyddwch} i ni gymwyſdꝛa: treuthwch i ni weniaith, \lhl{cenfyddwch} i ni ſiomedigaeth.}
		{ʾăšɛr ʾå̄mrū l·å̄·\hlA{rōʾīm} lō \hlA{ṯirʾū} wə·l·a·\lhl{ḥōzīm} lō \lhl{ṯɛḥɛ̆zū}·lå̄nū nəḵōḥōṯ dabbərū·lå̄nū ḥălå̄qōṯ \lhl{ḥăzū} mahăṯallōṯ}
		{Which say to the \hlA{seers}, \hlA{See} not; and to the \lhl{prophets}, \lhl{Prophesy} not unto us right things, speak unto us smooth things, \lhl{prophesy} deceits:}
	\end{example}
\end{frame}



\subsubsection{\bh{ṣå̄p̄å̄}}

\begin{frame}{\bh{ṣå̄p̄å̄} ‘to watch’}
	\begin{center}
		$$
		\mtext{\bh{ṣå̄p̄å̄}} \transto
		\begin{dcases}
			\mtext{\C{\hl{edrych}} (6 occurances)}\\
			\mtext{\C{disgwil} (3 occ.)}\\
			\mtext{\C{gwilio} (2 occ.)}\\
			\mtext{\C{craffu} (2 occ.)}\\
			\mtext{\C{canfod} (2 occ.)}\\
		\end{dcases}
		$$

		$$
		\mtext{\bh{ṣōp̄ɛ}} \transto
		\begin{dcases}
			\mtext{\C{\hl{gwiliedudd}} (12 occurances)}\\
			\mtext{\C{gwili-wr} (5 occ.)}\\
			\mtext{\small\C{disgwil-wr} (1 occ.)}
		\end{dcases}
		$$
	\end{center}
\end{frame}


\begin{frame}{\ex \bh{ṣå̄p̄å̄} ‘to watch’ → \C{edrych}}
	\begin{example}{Gen.}{31}{49}{}{}
		\quoling
		{\hlA{וְהַמִּצְפָּה֙} אֲשֶׁ֣ר אָמַ֔ר \lhl{יִ֥צֶף} יְהוָ֖ה בֵּינִ֣י וּבֵינֶ֑ךָ כִּ֥י נִסָּתֵ֖ר אִ֥ישׁ מֵרֵעֵֽהוּ׃}
		{A \hlA{Miſpah} hefyd, o blegit efe a ddywedaſe, \lhl{edꝛyched} yꝛ Arglwydd rhwng o fi a thithe, pan fôm ni bob vn o olwg ei gilydd.}
		{wə·ham·\hlA{miṣpå̄} ʾăšɛr ʾå̄mar \lhl{yīṣɛp̄} YHWH bēnī ū·ḇēnɛḵå̄ kī nisså̄ṯēr ʾīš mē·rēʿēhū}
		{And \hlA{Mizpah}; for he said, The {\LORD} \lhl{watch} between me and thee, when we are absent one from another.}
	\end{example}
\end{frame}


\begin{frame}{\ex \bh{ṣå̄p̄å̄} ‘to watch’ → \C{disgwil}}
	\begin{example}{Lam.}{4}{17}{}{}
		\quoling
		{*עודינה **עוֹדֵ֙ינוּ֙ תִּכְלֶ֣ינָה עֵינֵ֔ינוּ אֶל־עֶזְרָתֵ֖נוּ הָ֑בֶל \lhl{בְּצִפִּיָּתֵ֣נוּ} \lhl{צִפִּ֔ינוּ} אֶל־גּ֖וֹי לֹ֥א יוֹשִֽׁעַ׃}
		{Ein llygaid a ballaſant tra yꝛ oeddem ni yn \lhl{diſgwil} am ein cynnoꝛthwy ofer: \lhl{diſgwil} yꝛ oeddem ni yn vnic wꝛth genhedlaeth yꝛ hon nid achube.}
		{ʿōḏēnū tiḵlɛnå̄ ʿēnēnū ʾɛl·ʿɛzrå̄ṯēnū hå̄ḇɛl bə·\lhl{ṣippiyyå̄ṯēnū} \lhl{ṣippīnū} ʾɛl·gōy lō yōšīaʿ}
		{As for us, our eyes as yet failed for our vain help: in our watching we have watched for a nation \kjvit{that} could not save us.}
	\end{example}
\end{frame}


\begin{frame}{\ex \bh{ṣōp̄ɛ} ‘a watchman’ → \C{gwiliedudd}}
	\begin{example}{2 Sam.}{18}{24}{}{}
		\quoling
		{וְדָוִ֥ד יוֹשֵׁ֖ב בֵּין־שְׁנֵ֣י הַשְּׁעָרִ֑ים וַיֵּ֨לֶךְ \lhl{הַצֹּפֶ֜ה} אֶל־גַּ֤ג הַשַּׁ֙עַר֙ אֶל־הַ֣חוֹמָ֔ה וַיִּשָּׂ֤א אֶת־עֵינָיו֙ וַיַּ֔רְא וְהִנֵּה־אִ֖ישׁ רָ֥ץ לְבַדּֽוֹ׃}
		{A Dafydd oedd yn eiſtedd rhwng y ddau boꝛth: a’r \lhl{gwiliedudd} aeth ar nēn y poꝛth ar y mûr, ac a dderchafodd ei lygaid, ac a edꝛychodd, ac wele ŵꝛ yn rhedeg ei hunan.}
		{wə·ḏå̄wīḏ yōšēḇ bēn·šənē haš·šəʿå̄rīm way·yēlɛḵ haṣ·\lhl{ṣōp̄ɛ} ʾɛl·gaḡ haš·šaʿar ʾɛl·ha·ḥōmå̄ way·yiśśå̄ ʾɛṯ·ʿēnå̄w way·yar wə·hinnē·ʾīš rå̄ṣ lə·ḇaddō}
		{And David sat between the two gates: and the \lhl{watchman} went up to the roof over the gate unto the wall, and lifted up his eyes, and looked, and behold a man running alone.}
	\end{example}
\end{frame}



\subsubsection{\bh{šå̄r}}

\begin{frame}{\ex \bh{šå̄r} ‘to behold’ → \C{cael}~+ \C{gweled}}
	\begin{example}{Job}{24}{15}{}{}
		\quoling
		{וְעֵ֤ין נֹאֵ֨ף ׀ שָׁ֤מְרָֽה נֶ֣שֶׁף לֵ֭אמֹר לֹא־\lhl{תְשׁוּרֵ֣נִי} עָ֑יִן וְסֵ֖תֶר פָּנִ֣ים יָשִֽׂים׃}
		{A llygad y godineb-wꝛ ſydd yn gwilied y cyfnos gan ddywedyd, ni \hlA{chaiff} llygad fyng-\lhl{weled}: ac efe a eſyd hûg ar ei wyneb.}
		{wə-ʿēn nōʾēp̄ šå̄mrå̄ nɛšɛp̄ lē-mōr lō-\lhl{ṯəšūrēnī} ʿå̄yin wə-sēṯɛr på̄nīm yå̄śīm}
		{The eye also of the adulterer waiteth for the twilight, saying, No eye shall \lhl{see} me: and disguiseth his face.}
	\end{example}
\end{frame}

\subsubsection{\bh{hišqīp̄}~/ \bh{nišqå̄p̄}}

\begin{frame}{\bh{hišqīp̄}~/ \bh{nišqå̄p̄} ‘to view, to look from above’}
	\begin{center}
		$$
		\mtext{\bh{hišqīp̄} (12 occurances)} \transto
		\C{edrych}
		$$

		$$
		\mtext{\bh{nišqå̄p̄} (10 occ.)} \transto
		\begin{dcases}
			\mtext{\C{\hl{edrych}} (7 occ.)}\\
			\mtext{\C{gweled} (\gram{impers}) (3 occ.)}\\
		\end{dcases}
		$$
	\end{center}
\end{frame}


\begin{frame}{\ex \bh{hišqīp̄} ‘to view, to look from above’ → \C{edrych}}
	\begin{example}{2 Kings}{9}{30}{}{}
		\quoling
		{וַיָּב֥וֹא יֵה֖וּא יִזְרְעֶ֑אלָה וְאִיזֶ֣בֶל שָׁמְעָ֗ה וַתָּ֨שֶׂם בַּפּ֤וּךְ עֵינֶ֙יהָ֙ וַתֵּ֣יטֶב אֶת־רֹאשָׁ֔הּ \lhl{וַתַּשְׁקֵ֖ף} בְּעַ֥ד הַחַלּֽוֹן׃}
		{Pan ddaeth Iehu i Iezrahel: yna Iezabel a glybu [hynny,] ac a golurodd ei hwyneb, ac a wiſcodd yn wŷch am ei phenn, ac a \lhl{edꝛychodd} dꝛwy ’r ffeneſtr.}
		{way·yå̄ḇō yēhū yizrəʿɛl·å̄ wə·ʾīzɛḇɛl šå̄mʿå̄ wat·tå̄śɛm b·ap·pūḵ ʿēnɛhå̄ wat·tēṭɛḇ ʾɛṯ·rōšå̄h wat·\lhl{tašqēp̄} bəʿaḏ ha·ḥallōn}
		{And when Jehu was come to Jezreel, Jezebel heard \kjvit{of it}; and she painted her face, and tired her head, and \lhl{looked out} at a window.}
	\end{example}
\end{frame}


\begin{frame}{\ex \bh{nišqå̄p̄} → \C{edrych}}
	\begin{example}{Prov.}{7}{6}{}{}
		\quoling
		{כִּ֭י בְּחַלּ֣וֹן בֵּיתִ֑י בְּעַ֖ד אֶשְׁנַבִּ֣י \lhl{נִשְׁקָֽפְתִּי}׃}
		{Canys a mi yn ffeneſtr fy nhŷ, mi a \lhl{edꝛychais} dꝛwy fy ffeneſtr.}
		{kī bə·ḥallōn bēṯī bəʿaḏ ʾɛšnabbī \lhl{nišqå̄p̄tī}}
		{For at the window of my house I \lhl{looked} through my casement,}
	\end{example}
\end{frame}


\begin{frame}{\ex \bh{nišqå̄p̄ ʿal·pənē} → \C{edrych tua}}
	\begin{example}{Num.}{21}{20}{}{}
		\quoling
		{וּמִבָּמ֗וֹת הַגַּיְא֙ אֲשֶׁר֙ בִּשְׂדֵ֣ה מוֹאָ֔ב רֹ֖אשׁ הַפִּסְגָּ֑ה \lhl{וְנִשְׁקָ֖פָה} עַל־פְּנֵ֥י הַיְשִׁימֹֽן׃}
		{Ac o Bamoth [yn] y dyffryn, yꝛ hwn [ſydd] yng-wlad Moab, i ben y bꝛyn, yꝛ hwn ſydd yn \lhl{edꝛych} tua yꝛ diffaethwch.}
		{ū·mib·bå̄mōṯ hag·gay ʾăšɛr bi·śḏē mōʾå̄ḇ rōš hap·pisgå̄ wə·\lhl{nišqå̄p̄å̄} ʿal·pənē ha·yšīmōn}
		{And from Bamoth \lhl{in} the valley, that \lhl{is} in the country of Moab, to the top of Pisgah, which looketh toward Jeshimon.}
	\end{example}
\end{frame}


\begin{frame}{\ex \bh{nišqå̄p̄ ʿal·pənē} → \C{gwelir tua}}
	\begin{example}{Num.}{23}{28}{}{}
		\quoling
		{וַיִּקַּ֥ח בָּלָ֖ק אֶת־בִּלְעָ֑ם רֹ֣אשׁ הַפְּע֔וֹר \lhl{הַנִּשְׁקָ֖ף} עַל־פְּנֵ֥י הַיְשִׁימֹֽן׃}
		{A Balac a gymmerodd Balaam i ben Peoꝛ, yꝛ hwn a \lhl{welir} tua ’r diffaethwch.}
		{way·yiqqaḥ bå̄lå̄q ʾɛṯ·bilʿå̄m rōš hap·pəʿōr han·\lhl{nišqå̄p̄} ʿal·pənē ha·yšīmōn}
		{And Balak brought Balaam unto the top of Peor, that \lhl{looketh} toward Jeshimon.}
	\end{example}
\end{frame}



%%%\subsubsection{\bh{på̄qaḥ}}
%%%\subsubsection{\bh{šå̄zap̄}}



\subsubsection{\bh{hišgīaḥ}}

\begin{frame}{\ex \bh{hišgīaḥ} → \C{edrych}}
	\begin{example}{Song}{2}{9}{}{}
		\quoling
		{דּוֹמֶ֤ה דוֹדִי֙ לִצְבִ֔י א֖וֹ לְעֹ֣פֶר הָֽאַיָּלִ֑ים הִנֵּה־זֶ֤ה עוֹמֵד֙ אַחַ֣ר כָּתְלֵ֔נוּ \lhl{מַשְׁגִּ֙יחַ֙} מִן־הַֽחֲלֹּנ֔וֹת מֵצִ֖יץ מִן־הַֽחֲרַכִּֽים׃}
		{Tebyg yw fy annwylyd i iwꝛch neu lwdn hydd, wele ef yn ſefyll dann ein pared yn \lhl{edꝛych} trwy ’r ffeneſtri, [ac] yn ymddangos trwy ’r pilêrau.}
		{dōmɛ ḏōḏī li-ṣḇī ʾō lə-ʿōp̄ɛr hå̄-ʾayyå̄līm hinnē-zɛ ʿōmēḏ ʾaḥar kå̄ṯlēnū \lhl{mašgīaḥ} min-ha-ḥallōnōṯ mēṣīṣ min-ha-ḥărakkīm}
		{My beloved is like a roe or a young hart: behold, he standeth behind our wall, he \lhl{looketh} forth at the windows, shewing himself through the lattice.}
	\end{example}
\end{frame}



\subsubsection{Conclusion}

\begin{frame}{Types of translation relationships}
	\begin{itemize}
		\item Simple, no or very little variance:
			\bh{hɛrʾå̄}, \bh{nirʾå̄}, \bh{hibbīṭ}, \bh{ḥå̄zå̄}, \bh{hišgīaḥ}, \bh{rōʾɛ}, \bh{ḥōzɛ}.
		\item Two main options:
			\bh{rå̄ʾå̄}, \bh{på̄nå̄}.
		\item More variance:
			\bh{ṣå̄p̄å̄}.
	\end{itemize}
\end{frame}
