\subsection{Taste}

\sepslide{Taste}



\subsubsection{\bh{ṭå̄ʿam}}

\begin{frame}{\bh{ṭå̄ʿam} ‘to taste’}
	\begin{center}
		$$
		\mtext{\bh{ṭå̄ʿam}} \transto
		\begin{dcases}
			\mtext{\C{\hl{archwaithu}}\quad \small (7 occurances)}\\
			\mtext{\C{profi}\quad \small (2 occurances)}\\
			\mtext{\C{gweled}\quad \small (one occurance)}
		\end{dcases}
		$$
	\end{center}
\end{frame}

\begin{frame}{\ex \bh{ṭå̄ʿam} ‘to taste’ → \C{archwaithu}}
	\begin{example}{1 Sam.}{14}{29}{}{}
		\quoling
		{[…] רְאוּ־נָא֙ כִּֽי־אֹ֣רוּ עֵינַ֔י כִּ֣י \lhl{טָעַ֔מְתִּי} מְעַ֖ט דְּבַ֥שׁ הַזֶּֽה׃}
		{[…], gwelwch yn awꝛ fel y goleuodd fy llygaid o herwydd i mi \lhl{archwaithu} y chydig o’ꝛ mêl hwn.}
		{[…] rəʾū-nå̄ kī-ʾōrū ʿēnay kī \lhl{ṭå̄ʿamtī} məʿaṭ dəḇaš haz-zɛ}
		{[…]: see, I pray you, how mine eyes have been enlightened, because I \lhl{tasted} a little of this honey.}
	\end{example}
\end{frame}

\begin{frame}{\ex \bh{ṭå̄ʿam} ‘to taste’ → \C{profi}}
	\begin{example}{Jon.}{3}{7}{}{}
		\quoling
		{[…] הָאָדָ֨ם וְהַבְּהֵמָ֜ה הַבָּקָ֣ר וְהַצֹּ֗אן אַֽל־\lhl{יִטְעֲמוּ֙} מְא֔וּמָה אַ֨ל־יִרְע֔וּ וּמַ֖יִם אַל־יִשְׁתּֽוּ׃}
		{[…]: dŷn, ac anifail, eidion, a dafad, na \lhl{phꝛofant} ddim, na phoꝛant, ac nac yſant ddwfr.}
		{[…] hå̄-ʾå̄ḏå̄m wə-hab-bəhēmå̄ hab-bå̄qå̄r wə-haṣ-ṣōn ʾal-\lhl{yiṭʿămū} məʾūmå̄ ʾal-yirʿū ū-mayim ʾal-yištū}
		{[…], Let neither man nor beast, herd nor flock, \lhl{taste} any thing: let them not feed, nor drink water:}
	\end{example}
\end{frame}

\begin{frame}{\ex \bh{ṭå̄ʿam} ‘to taste’ → \C{gweled}}
	\begin{example}{Prov.}{31}{18}{}{}
		\quoling
		{\lhl{טָ֭עֲמָה} כִּי־ט֣וֹב סַחְרָ֑הּ לֹֽא־יִכְבֶּ֖ה *בליל **בַלַּ֣יְלָה נֵרָֽהּ׃}
		{Os hi a \lhl{wêl} fôd ei marſiandiaeth yn fuddiol, ni ddiffydd ei chanwyll ar hŷd y nôs.}
		{\lhl{ṭå̄ʿămå̄} kī·ṭōḇ saḥrå̄h lō·yiḵbɛ ḇ·al·laylå̄ nērå̄h}
		{She \lhl{perceiveth} that her merchandise \kjvit{is} good: her candle goeth not out by night.}
	\end{example}
\end{frame}



\subsubsection{\bh{ṭaʿam}}

\begin{frame}{\bh{ṭaʿam} ‘taste’ (noun)}
	\begin{center}
		$$
		\text{\bh{ṭaʿam}}
		\begin{dcases}
			\mtext{sensory taste} \transto \C{blâs}\\
			\mtext{figurative meaning} \transto
			\begin{dcases}
				\C{deall}\\
				\C{rhesum}\\
				\C{synnwyr}\\
				\C{cyngor}\\
				\C{ymadrodd}\\
				\C{agwedd}\\
				\mtext{\C{(dꝛwy) oꝛchymyn}}
			\end{dcases}
		\end{dcases}
		$$
	\end{center}
\end{frame}



\begin{frame}{\ex \bh{ṭaʿam} ‘taste’, in the literal sense}
	\begin{example}{Ex.}{16}{31}{}{}
		\quoling
		{וַיִּקְרְא֧וּ בֵֽית־יִשְׂרָאֵ֛ל אֶת־שְׁמ֖וֹ מָ֑ן וְה֗וּא כְּזֶ֤רַע גַּד֙ לָבָ֔ן \lhl{וְטַעְמ֖וֹ} כְּצַפִּיחִ֥ת בִּדְבָֽשׁ׃}
		{A thŷ Iſrael a alwaſant ei henw ef Manna: ac efe oedd fel hâd Coꝛiander, yn wynn, ai \lhl{flâs} fel afrllad o fêl.}
		{way·yiqrəʾū ḇēṯ·yiśrå̄ʾēl ʾɛṯ·šəmō må̄n wə·hū kə·zɛraʿ gaḏ lå̄ḇå̄n wə·\lhl{ṭaʿmō} kə·ṣappīḥīṯ bi·ḏḇå̄š}
		{And the house of Israel called the name thereof Manna: and it was like coriander seed, white; and the \lhl{taste} of it was like wafers made with honey.}
	\end{example}
\end{frame}



\begin{frame}{\ex \bh{ṭaʿam} ‘taste’, in a figurative sense}
	\begin{example}{Ps.}{119}{66}{}{}
		\quoling
		{ט֤וּב \lhl{טַ֣עַם} וָדַ֣עַת לַמְּדֵ֑נִי כִּ֖י בְמִצְוֺתֶ֣יךָ הֶאֱמָֽנְתִּי׃}
		{Dyſc i mi iawn \lhl{ddeall}, a gŵybodaeth, o herwydd yn dy oꝛchymynnion di y credais.}
		{ṭūḇ \lhl{ṭaʿam} wå̄·ḏaʿaṯ lamməḏēnī kī ḇə·miṣwōṯɛḵå̄ hɛʾɛ̆må̄ntī}
		{Teach me good \lhl{judgment} and knowledge: for I have believed thy commandments.}
	\end{example}
\end{frame}



\begin{frame}{\ex \bh{miṭ-ṭaʿam} ‘by decree’}
\begin{example}{Jon.}{3}{7}{}{}
	\quoling
	{וַיַּזְעֵ֗ק וַיֹּ֙אמֶר֙ בְּנִֽינְוֵ֔ה \lhl{מִטַּ֧עַם} הַמֶּ֛לֶךְ וּגְדֹלָ֖יו לֵאמֹ֑ר […]}
	{Yna y cyhoeddwyd, ac y dywedwyd yn Ninife \lhl{dꝛwy oꝛchymyn} y bꝛenin ai bendefigion, gan ddywedyd: […]}
	{way·yazʿēq way·yōmɛr bə·nīnwē \lhl{miṭ·ṭaʿam} ham·mɛlɛḵ ū·ḡḏōlå̄w lē·mōr […]}
	{And he caused \textit{it} to be proclaimed and published through Nineveh \lhl{by the decree} of the king and his nobles, saying, […]}
\end{example}
\end{frame}



\subsubsection{\bh{maṭʿammīm}}

\begin{frame}{\ex \bh{maṭʿammīm} ‘delicacies’ → \C{dainteithion}}
	\begin{example}{Prov.}{23}{3}{}{}
		\quoling
		{אַל־תִּ֭תְאָו‪[Q]‬‪[q]‬ \lhl{לְמַטְעַמּוֹתָ֑יו} וְ֝ה֗וּא לֶ֣חֶם כְּזָבִֽים׃}
		{Na ddeiſyf ei \lhl{ddanteithion} ef: canys twyllodꝛus yw bwyd.}
		{ʾal·tiṯʾå̄w lə·\lhl{maṭʿammōṯå̄w} wə·hū lɛḥɛm kəzå̄ḇīm}
		{Be not desirous of his \lhl{dainties}: for they are deceitful meat.}
	\end{example}
\end{frame}
